---
title: "Analyzing Non-Voters in the US: A Tidying and Transformation Exercise in R"
author: "Mohammed Rahman"
date: "2023-03-06"
output: html_document
---

<!-- https://rpubs.com/mrcuny/analyzing-non-voters-in-the-us -->

[GitHub Link to Rmd file](https://github.com/mrcuny/analyzing-non-voters-in-the-us/blob/main/analyzing_non_voters_in_the_us.Rmd)

----

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This analysis is based on a dataset behind the story [Why Many Americans Don't Vote](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/) from FiveThirtyEight that explores why some Americans do not vote in presidential elections.

The original dataset contains information on a variety of demographic, socioeconomic, and attitudinal factors that contribute to non-voting. 

In this analysis, we will be selecting a subset of variables to explore the relationships between non-voting and key demographic and socioeconomic factors.

## Data Cleaning and Transformation

First, let's load the libraries we will be using:

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
```

Next, we will load the original dataset into R:

```{r warning=FALSE, message=FALSE}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv"
nonvoters <- read_csv(url)
```

This dataset has `r ncol(nonvoters)` columns and `r nrow(nonvoters)` rows.

Here are the column names:

```{r}
colnames(nonvoters)
```

### Select relevant columns

We will select a subset of variables from the dataset that are most relevant to our analysis. Specifically, we will focus on the following variables:

- ppage: Age of respondent
- educ: Highest educational attainment category
- race: Race of respondent, census categories
- gender: Gender of respondent
- income_cat: Household income category of respondent
- voter_category: Voter category of respondent: always, sporadic or rarely/never

```{r}
nonvoters <- nonvoters %>%
  select(ppage, educ, race, gender, income_cat, voter_category)
```

Let's view the first `10` rows:

```{r}
head(nonvoters, 10)
```


### Rename columns

Let's use better and clear column names, and get rid of unnecessary abbreviations:

```{r}
nonvoters <- nonvoters %>%
  rename(
    age = ppage,
    highest_education_level = educ,
    income_category = income_cat
  )
```

The renamed columns now look like this:

```{r}
head(nonvoters, 10)
```


## Conclusions

In conclusion, the non-voters dataset provides valuable insights into the demographics and attitudes of non-voters in the United States. By tidying the dataset, we were able to create a more useful and manageable subset of the data that includes only the relevant columns. We also renamed some of the columns to make them more interpretable.

To extend and verify the work from the selected article, one could conduct more detailed analyses to investigate the reasons behind non-voting among different demographic groups. This could include conducting statistical tests to determine if there are significant differences in the reasons for non-voting between different age, race, or income groups. Additionally, it would be interesting to explore the effectiveness of various voter outreach and mobilization strategies, such as door-to-door canvassing, phone banking, or social media campaigns. Such analyses could help to identify effective strategies for increasing voter turnout in future elections.
